The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-04 manuscript/manuscript.Rnw
include hormone/igf/galectin heatmap figure

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -1, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")

@@

>>>>

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
options(scipen = -1, digits = 2)
library("data.table")
library("ggplot2")
library("reshape2")
library("xtable")
library("grid")
library("gridExtra")
library("ape")
library("ggtree")
library("topGO")
library("org.Hs.eg.db")
library("scales")

@@

\begin{figure}
<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

gene.expression.nozero <- function(genes) {
    temp <-
        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
                             human_name~species,
                             value.var="FPKM",
                             fun.aggregate=function(x){x[1]}))
    temp <-
        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
    return(temp)
}

### hormone genes
hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
                   "GH2","GHRH","HSD3B1","NPPB","NPPC","PRL","RLN1")
hormone.expression.nozero <-
    gene.expression.nozero(hormone.genes)

hor.exp.long <- melt(log10(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]

### galectin genes
galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))

galectin.expression.nozero <-
    gene.expression.nozero(galectin.genes)

gal.exp.long <- melt(log10(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
colnames(gal.exp.long)[1:3] <- c("gene","species","fpkm")

### IGF related genes

igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))

igfrelated.expression.nozero <-
    gene.expression.nozero(igfrelated.genes)

igf.exp.long <- melt(log10(igfrelated.expression.nozero+1)[rev(1:nrow(igfrelated.expression.nozero)),])
colnames(igf.exp.long)[1:3] <- c("gene","species","fpkm")

igf.exp.long$type <- "IGF Related"
gal.exp.long$type <- "Galectin"
hor.exp.long$type <- "Hormone"

### scalerange <- range(c(igf.exp.long$fpkm,gal.exp.long$fpkm,hor.exp.long$fpkm),na.rm=TRUE)
### gradientends <- scalerange + rep(c(0,100,200), each=2)
### colorends <- c("white", "red", "white", "green", "white", "blue")
### 
### gal.exp.long$fpkm <- gal.exp.long$fpkm+100
### hor.exp.long$fpkm <- hor.exp.long$fpkm+200

combined.long <-
    rbind(igf.exp.long,
          gal.exp.long,
          hor.exp.long)
combined.long$species <-
    factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(combined.long$species)),
           levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(names(species.names))))

common.plot.options <- 
    list(geom_tile(aes(fill = fpkm), colour = "white"),
##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
         scale_fill_gradient(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "grey50",
                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
         scale_x_discrete("", expand = c(0, 0)),
         scale_y_discrete("", expand = c(0, 0)),
         theme_grey(base_size = 9),
         theme(legend.position = "top",
               axis.ticks = element_blank(), 
               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
         facet_wrap(~type,scale="free_y",ncol=1)
         )
print(ggplot(combined.long, aes(y=gene, x=species))
      + common.plot.options
      )
@ 
\caption{Expression of Hormone Genes, Galectins, and IGF related genes}
\label{fig:hormone_genes}
\end{figure}
